Modal Abstractions for Virtualizing Memory Addresses

### 2025-08-17 background.tex
various wordsmithing to reduce overage; guessing at camera-ready options, have cut overage roughly in half; next target is sections 4.2/4.3 which has a large gap after figure 7

The entries \add{(e.g., L4 Entry in Figure \ref{fig:pagetables})} of each table are 64 bits wide,
but each points to a physical address aligned to 4KB (4096 byte) boundaries, which leaves 12 bits to spare to 
control a validity bit (called the \emph{present} bit), a read-write bit (which permits write access through the entry if and only if it is set),
and a range of additional bits which can be used to control caching, write-through, and more.
This paper will only consider the present bit (0).

The page tables are managed by the \replace{OS, typically by a}{OS kernel's} \emph{virtual memory manager} (VMM).\footnote{Not to be confused with Virtual 
Machine Monitor. We focus on non-hypervisor scenarios, but hardware virtualization extensions for both 
x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its 
(virtualized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
Typically each process has its own 
page tables, which the OS registers with the CPU by storing the (page-aligned)
physical address of the root of the page table tree (the start of the L4 table shown as $\mathsf{root}$ in Figure \ref{fig:pagetables}) in a specific register (\texttt{cr3}) as part of switching to a new process. 
Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) 
is how the OS ensures memory isolation between processes.

If an instruction is executed that accesses a virtual address that either has no mapping, or does not have a mapping permitting 
the kind of access that was performed (e.g., the instruction was a memory write, but the relevant address range was marked read-only in
the relevant page table entry), the hardware triggers a \emph{page fault}, transferring control to a \emph{page fault handler} registered 
with the hardware by the OS, allowing it to take corrective action.
If no mapping was supposed to exist, this is a program bug (e.g., dereferencing virtual address 0 / NULL)
and the faulting program should be terminated. But this can also be used for
\replace{
specialized functionality and optimizations, such as \emph{paging} (saving room in physical RAM and deferring
unnecessary IO by only reading program code from disk when it is accessed, or even swapping memory that has not
recently been accessed to disk, to read back in when a page fault indicates access).
}{
  other purposes such as demand paging to save on IO and better-manage physical memory~\cite{doeppner2010operating}.
}

@@

Virtual memory concerns propagate to the OS scheduler, which deals with multiple 
address spaces, so must keep track of which virtual addresses are valid (and in what way) in which address spaces. 
Some virtual addresses are valid in only a single address space (e.g., a code address for a particular usermode 
process), while others are valid in all address spaces (e.g., kernel data structure pointers). 
The VMM must maintain some of these assumptions on behalf of the rest of the kernel, for example by guaranteeing that 
a certain range of virtual addresses (corresponding to the kernel's code and data) are valid in every address space.

@@

\iris includes two forms of implication. The magic wand operator $\wand$ is an affine implication:
$A\wand B$ describes a resource which, if combined with a resource satisfying $A$, will satisfy $B$.
Notably, this implication involves no changes to ghost state. \iris, building on the Views framework~\cite{Dinsdale-Young2013Views},
also includes a \emph{view shift} operator $\vs$ which models updates to ghost state: $A \vs B$
means resources satisfying $A$ may be transformed into resources satisfying $B$, intuitively by updating only ghost state.\footnote{\iris
experts may note that this is \emph{technically} mildly misleading given how \iris's update modalities and weakest precondition
are defined, but
this is adequate intuition for non-experts in \iris to follow their use.
}
View shifts are essentially logical entailment plus ghost updates.
\looseness=-1

>>>>

The entries \add{(e.g., L4 Entry in Figure \ref{fig:pagetables})} of each table are 64 bits wide,
but each points to a physical address aligned to 4KB (4096 byte) boundaries, leaving 12 bits to spare to 
control a validity bit (called the \emph{present} bit), a read-write bit (enabling write access),
and a range of additional bits which can be used to control caching, write-through, and more.
This paper will only consider the present bit (0).
\looseness=-1

The page tables are managed by the \replace{OS, typically by a}{OS kernel's} \emph{virtual memory manager} (VMM).\footnote{Not to be confused with Virtual 
Machine Monitor. We focus on non-hypervisor scenarios, but hardware virtualization extensions for both 
x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its 
(virtualized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
Typically each process has its own 
page tables, which the OS registers with the CPU by storing the (page-aligned)
physical address of the root of the page table tree (the start of the L4 table shown as $\mathsf{root}$ in Figure \ref{fig:pagetables}) in a specific register (\texttt{cr3}) as part of switching to a new process. 
Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) 
is how the OS ensures memory isolation between processes.
If an instruction is executed that accesses a virtual address that either has no mapping, or does not have a mapping permitting 
the kind of access that was performed (e.g., the instruction was a memory write, but the relevant address range was marked read-only in
the relevant page table entry), the hardware triggers a \emph{page fault}, transferring control to a \emph{page fault handler} registered 
with the hardware by the OS, allowing it to take corrective action.
If no mapping was supposed to exist, this is a program bug (e.g., dereferencing virtual address 0 / NULL)
and the faulting program should be terminated. But this can also be used for
other purposes such as demand paging to save on IO and better-manage physical memory~\cite{doeppner2010operating}.
\looseness=-1

@@

Virtual memory affects the OS scheduler, which deals with multiple 
address spaces, so must track which virtual addresses are valid in which address spaces. 
Some virtual addresses are valid in only a single address space (e.g., a code address for a particular usermode 
process), while others are valid in all address spaces (e.g., kernel data structure pointers). 
The VMM must maintain some of these assumptions on behalf of the rest of the kernel, for example by guaranteeing that 
a certain range of virtual addresses (corresponding to the kernel's code and data) are valid in every address space.
\looseness=-1

@@

\iris includes two forms of implication. The magic wand operator $\wand$ is an affine implication:
$A\wand B$ describes a resource which, if combined with a resource satisfying $A$, will satisfy $B$.
Notably, this implication involves no changes to ghost state. \iris, building on the Views framework~\cite{Dinsdale-Young2013Views},
also includes a \emph{view shift} operator $\vs$ which models updates to ghost state: $A \vs B$
means resources satisfying $A$ may be transformed into resources satisfying $B$, intuitively by updating only ghost state
(a slight simplification of \iris's update modalities, but adequate intuition for non-\iris-experts).
% .\footnote{\iris
% experts may note that this is \emph{technically} mildly misleading given how \iris's update modalities and weakest precondition
% are defined, but
% this is adequate intuition for non-experts in \iris to follow their use.
% }
View shifts are essentially logical entailment plus ghost updates.
\looseness=-1

### 2025-08-17 syntaxsemantics.tex
add keywords and CCS; remove all uses of \del
various wordsmithing to reduce overage; guessing at camera-ready options, have cut overage roughly in half; next target is sections 4.2/4.3 which has a large gap after figure 7

Programs in our logic are instruction sequences \instrs, which are formed by either a basic instruction \iskip, or prefixing an existing instruction
sequence with an additional instruction (\iseq\instr\instrs).
We model (and later, give program logic rules for) instructions for \del{basic} register \replace{moves}{loads and stores}, and reading and writing memory.
The latter require page table walks.
% Figure \ref{fig:coq_addr_translation} gives a slightly simplified version of our address translation code in monadic style:
% starting from the root page table address \textsf{rtv}, the virtual address \textsf{w} is either resolved to a physical address or an error is returned.
% The full set of possible errors is not relevant to this paper; our logic is tailored for kernel code which should not fault,
% so our proof rules guarantee no memory failures occur --- that the page table walk for any dereferenced address will succeed.

% Page-table walk shown in Figure \ref{fig:pagetables} is realized with physical memory \texttt{load}s and \textsf{store}s for the entries in the 
% page-tables which are simply modelled as lookup and updates to an instance of a memory map with 64 bit entries. In the address translation,
% we see a physical memory $\mathsf{load}$ (i.e. physical memory map ($\sigma.\mathcal{M}$) for each level) for each level of page-table-walk 
% concluded with alignment of the returned address for each \textit{successful} page-table-walk traversal which is what our reasoning principles 
% only consider.

The most important instructions that we model are memory-accessing variants of the x86-64 \lstinline|mov| instruction, which we format in Intel syntax
(destination on the left, source on the right).
Thus, a store to memory is $\textsf{mov}~[r_m]~r_r$, and a load from memory is $\textsf{mov}~r_r~[r_m]$.
% \[
% \begin{array}{l}
%   \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad
%   \textsf{mov}~[r_m]~r_r \\
%   \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 
%   \textsf{mov}~r_r~[r_m]
%   \end{array}
%   \]
Operationally, each first translates the virtual address stored in the register $r_m$ to a physical address \add{(a page table walk)},
then either updates the memory at that physical location with the contents of $r_r$ (for store)
or updates $r_r$ with the contents of that memory (for load).
% The semantics of each is realized by first translating the virtual-memory address of a memory location stored in $r_m$ to a physical memory location
% per the description in Section \ref{sec:backgroundonmachinemodel},
% then either updating those memory contents with the contents of register $r_r$ (\TirNameStyle{WriteToMemFromReg}) 
% or loading the value at that physical memory location into the register $r_r$
% (\TirNameStyle{WriteToRegFromMem}). 
Our formalization includes
additional \lstinline|mov| variants (e.g., accessing memory at constant offsets from the base register, or moves between registers),
basic integer and bitwise operations (\lstinline|add|,
\lstinline|and|, bit shifts, etc.) with their effects on \lstinline|rflags|, jumps and some (not all) conditional jumps, \lstinline|call|, \lstinline|ret|, \lstinline|push|,
and \lstinline|pop|.
The semantics model updates to the instruction pointer, though we elide that register (\lstinline|rip|)
from specifications in the paper for clarity.
\looseness=-1

>>>>

Programs in our logic are instruction sequences \instrs, which are formed by either a basic instruction \iskip, or prefixing an existing instruction
sequence with an additional instruction (\iseq\instr\instrs).
We model (and later, give program logic rules for) instructions for register \replace{moves}{loads and stores}, and reading and writing memory.
The latter require page table walks.
The most important instructions that we model are memory-accessing variants of the x86-64 \lstinline|mov| instruction, which we format in Intel syntax
(destination on the left, source on the right).
Thus, a store to memory is $\textsf{mov}~[r_m]~r_r$, and a load from memory is $\textsf{mov}~r_r~[r_m]$.
Operationally, each first translates the virtual address stored in the register $r_m$ to a physical address \add{(a page table walk)},
then either updates the memory at that physical location with the contents of $r_r$ (for store)
or updates $r_r$ with the contents of that memory (for load).
Our formalization includes
additional \lstinline|mov| variants (e.g., accessing memory at constant offsets from the base register, or moves between registers),
basic integer and bitwise operations (\lstinline|add|,
\lstinline|and|, bit shifts, etc.) with their effects on \lstinline|rflags|, jumps and some (not all) conditional jumps, \lstinline|call|, \lstinline|ret|, \lstinline|push|,
and \lstinline|pop|.
% The semantics model updates to the instruction pointer, though we elide that register (\lstinline|rip|)
% from specifications in the paper for clarity.
\looseness=-1
